### 2025-11-17 cluster-analysis.qmd
20251111 - reorder
20251117 - fixes

An overview of approaches to cluster analysis in `R` is provided by @Kassambara2017a.
In this chapter, we focus on examples using model-based clustering with the `R` package `mclust` [@R-mclust; @Scrucca2023_packages], which uses Gaussian finite mixture modeling.
Model-based clustering assumes the data are generated by an underlying statistical model, and the goal is to recover the structure of that model.
Model-based clustering represents the data as coming from a mixture of probability distributions.
The various types of `mclust` models are provided here:
<https://mclust-org.github.io/mclust/reference/mclustModelNames.html>.

@@

@@

tiersQB_boostrap <- mclust::mclustBootstrapLRT(
  data = player_stats_seasonal_offense_recentQB$fantasyPoints,
  modelName = "V") # variable/unequal variance (for univariate data)

numTiersQB <- as.numeric(summary(tiersQB_boostrap)[,"Length"][1]) # or could specify the number of teams manually

tiersQB_boostrap
plot(
  tiersQB_boostrap,
  G = numTiersQB - 1)
```

@@

<!--
The model-based bootstrap clustering of Running Backs' fantasy points is unable to run due to an error:
-->

```{r}
tiersRB_boostrap <- mclust::mclustBootstrapLRT(
  data = player_stats_seasonal_offense_recentRB$fantasyPoints,
  modelName = "V") # variable/unequal variance (for univariate data)
```

<!--
Thus, we cannot use the following code, which would otherwise summarize the model results, specify the number of tiers, and plot model comparisons:
-->

```{r}
numTiersRB <- as.numeric(summary(tiersRB_boostrap)[,"Length"][1]) # or could specify the number of teams manually

tiersRB_boostrap
plot(
  tiersRB_boostrap,
  G = numTiersRB - 1)
```

@@

tiersWR_boostrap <- mclust::mclustBootstrapLRT(
  data = player_stats_seasonal_offense_recentWR$fantasyPoints,
  modelName = "V") # variable/unequal variance (for univariate data)

numTiersWR <- as.numeric(summary(tiersWR_boostrap)[,"Length"][1]) # or could specify the number of teams manually

tiersWR_boostrap
plot(
  tiersWR_boostrap,
  G = numTiersWR - 1)
```

@@

@@

tiersTE_boostrap <- mclust::mclustBootstrapLRT(
  data = player_stats_seasonal_offense_recentTE$fantasyPoints,
  modelName = "V") # variable/unequal variance (for univariate data)

numTiersTE <- as.numeric(summary(tiersTE_boostrap)[,"Length"][1]) # or could specify the number of teams manually

tiersTE_boostrap
plot(
  tiersTE_boostrap,
  G = numTiersTE - 1)
```

@@

wrTypes_boostrap <- mclust::mclustBootstrapLRT(
  data = merged_dataWRs_cluster %>% select(-gsis_id, -display_name),
  modelName = "EVE") # ellipsoidal with equal volume, variable shape, and equal orientation (for multivariate data)

wrTypes_boostrap
plot(
  wrTypes_boostrap,
  G = numTypesWR - 1)
```

>>>>

An overview of approaches to cluster analysis in `R` is provided by @Kassambara2017a.
In this chapter, we focus on examples using model-based clustering with the `R` package `mclust` [@R-mclust; @Scrucca2023_packages], which uses Gaussian finite mixture modeling.
Model-based clustering assumes the data are generated by an underlying statistical model, and the goal is to recover the structure of that model.
Model-based clustering represents the data as coming from a mixture of probability distributions.

@@

The various types of `mclust` models are provided here:
<https://mclust-org.github.io/mclust/reference/mclustModelNames.html>.

@@

tiersQB_bootstrap <- mclust::mclustBootstrapLRT(
  data = player_stats_seasonal_offense_recentQB$fantasyPoints,
  modelName = "V") # variable/unequal variance (for univariate data)

numTiersQB <- as.numeric(summary(tiersQB_bootstrap)[,"Length"][1]) # or could specify the number of teams manually

tiersQB_bootstrap
plot(
  tiersQB_bootstrap,
  G = numTiersQB - 1)
```

@@

<!--
The model-based bootstrap clustering of Running Backs' fantasy points is unable to run due to an error:
UPDATE: THE ERROR NO LONGER OCCURS
-->

```{r}
tiersRB_bootstrap <- mclust::mclustBootstrapLRT(
  data = player_stats_seasonal_offense_recentRB$fantasyPoints,
  modelName = "V") # variable/unequal variance (for univariate data)
```

<!--
Thus, we cannot use the following code, which would otherwise summarize the model results, specify the number of tiers, and plot model comparisons:
UPDATE: THE ERROR NO LONGER OCCURS
-->

```{r}
numTiersRB <- as.numeric(summary(tiersRB_bootstrap)[,"Length"][1]) # or could specify the number of teams manually

tiersRB_bootstrap
plot(
  tiersRB_bootstrap,
  G = numTiersRB - 1)
```

@@

tiersWR_bootstrap <- mclust::mclustBootstrapLRT(
  data = player_stats_seasonal_offense_recentWR$fantasyPoints,
  modelName = "V") # variable/unequal variance (for univariate data)

numTiersWR <- as.numeric(summary(tiersWR_bootstrap)[,"Length"][1]) # or could specify the number of teams manually

tiersWR_bootstrap
plot(
  tiersWR_bootstrap,
  G = numTiersWR - 1)
```

@@

###### Model-Based Clustering {#sec-clusterAnalysisExampleNumTiersTEsModel}

@@

tiersTE_bootstrap <- mclust::mclustBootstrapLRT(
  data = player_stats_seasonal_offense_recentTE$fantasyPoints,
  modelName = "V") # variable/unequal variance (for univariate data)

numTiersTE <- as.numeric(summary(tiersTE_bootstrap)[,"Length"][1]) # or could specify the number of teams manually

tiersTE_bootstrap
plot(
  tiersTE_bootstrap,
  G = numTiersTE - 1)
```

@@

wrTypes_bootstrap <- mclust::mclustBootstrapLRT(
  data = merged_dataWRs_cluster %>% select(-gsis_id, -display_name),
  modelName = "EVE") # ellipsoidal with equal volume, variable shape, and equal orientation (for multivariate data)

wrTypes_bootstrap
plot(
  wrTypes_bootstrap,
  G = numTypesWR - 1)
```
